---
title: "Solutions to dplyr introduction"
author: "Mark Dunning"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---


```{r}
library(dplyr)
library(stringr)

patients <- read.delim("patient-data.txt")
patients <- tbl_df(patients)
```


- Print all the columns between `Height` and `Grade_Level`

```{r}
select(patients, Height:Grade_Level)
```

- Print all the columns between `Height` and `Grade_Level`, but NOT `Pet`

```{r}
select(patients, Height:Grade_Level,Pet)
```

- Print the columns `Height` and `Weight`
    + try to do this without specifying the full names of the columns
    
```{r}
select(patients, contains("eight"))
select(patients, ends_with("eight"))
```

- (OPTIONAL)
- Print the columns in alphabetical order
- Print all the columns whose name is less than 4 characters in length

```{r}
select(patients, order(colnames(patients)))
select(patients, which(nchar(colnames(patients))<4))
```


- We want to calculate the Body Mass Index (BMI) for each of our patients
- $BMI = (Weight) / (Height^2)$ 
    + where Weight is measured in Kilograms, and Height in Metres
- Create a new BMI variable in the dataset
- A BMI of 25 is considered overweight, calculate a new variable to indicate which individuals are overweight
- For a follow-on study, we are interested in overweight smokers
    + clean the `Smokes` column to contain just `TRUE` or `FALSE` values
- How many candidates (Overweight and Smoker) do you have?
- (EXTRA) What other problems can you find in the data?


```{r}
patients_clean <- mutate(patients, Sex = factor(str_trim(Sex)))
patients_clean <- mutate(patients_clean, Height= as.numeric(str_replace_all(Height,pattern = "cm","")))
patients_clean <- mutate(patients_clean, Weight = as.numeric(str_replace_all(patients_clean$Weight,"kg","")))
patients_clean <- mutate(patients_clean, BMI = (Weight/(Height/100)^2), Overweight = BMI > 25)
patients_clean <- mutate(patients_clean, Smokes = str_replace_all(Smokes, "Yes", TRUE))
patients_clean <- mutate(patients_clean, Smokes = as.logical(str_replace_all(Smokes, "No", FALSE)))

```

