---
title: "Tidying and transforming data - live coding script"
author: "Matt Eldridge"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
---

## Introduction

In this session we will look at what makes a dataset 'tidy' and why this is important. We will
also look at how to perform some basic manipulations of a tabular dataset including:

* operations that will help clean up values within columns
* selecting a subset of columns of interest
* creating new variables (columns) based on existing variables

### Load the tidyverse

In this section we will use functions from a number of packages from the _tidyverse_,
namely **tidyr** for tidying data, **dplyr** for manipulating data frames and **stringr**
providing common string operations.

Load the core packages from the _tidyverse_.

```{r}
library(tidyverse)
```

_[**Instructors note:** draw attention to the packages that are loaded, relate these to this and other sections of the course.]_

#### Aside: function name conflicts (optional)

With the many hundreds of R packages available it is inevitable that there will
be functions with the same name being used by more than one package.

For example, filter function in dplyr  has masked filter function from the stats package that
already loaded as part of base R.

```{r eval = FALSE}
?filter
```

Navigate to each help section and show how the signatures differ and explain how using
the one you hadn't intended will likely cause your code to fail.

Use R studio's file completion to show that the filter function we get by default.

To use the `filter` function from the `stats` package we need to include the `stats`
namespace prefix as follows:

```{r eval = FALSE}
stats::filter(presidents, rep(1, 3))
```

### Data frames

In this course the data we will be working with will be structured within a type of R
object called a 'data frame'.

A data frame contains tabular data, like those we might be used to dealing with in spreadsheets,
e.g. with Excel, where the data can be thought of as having rows and columns. The values in each
column are all of the same type, e.g. all numbers (if quantitative) or all text strings (if qualitative).

Let's have a look at an example data frame that was loaded as part of the `tidyr`
package that we loaded when we called `library(tidyverse)`.

This dataset contains the number of TB cases documented by the World Health Organization
in Afghanistan, Brazil and China in 1999 and 2000.

```{r eval = FALSE}
table1
```

If we view this in the R console it is displayed as follows and described as a `tibble`.

```
# A tibble: 6 x 4
  country      year  cases population
  <chr>       <int>  <int>      <int>
1 Afghanistan  1999    745   19987071
2 Afghanistan  2000   2666   20595360
3 Brazil       1999  37737  172006362
4 Brazil       2000  80488  174504898
5 China        1999 212258 1272915272
6 China        2000 213766 1280428583
```

```{r}
class(table1)
```

This is a special kind of data frame with the additional `tbl_df` and `tbl` types.

Tibbles have some additional behaviour over plain data frames that might be useful.
The most obvious is that these get printed to the console in a more concise and
useful way, particularly when there are very many rows and/or columns. You can see
how the default behaviour of tibbles differs from normal data frames by looking at
the help page for the `tbl_df` class.

```{r eval = FALSE}
?tbl_df
```

## Tidy and untidy data

The **tidyr** package also contains some different representations of the TB dataset.
Each of these tables organizes the values in a different layout or format.

Let's take a look at the different representations.

`table2` contains separate rows for the number of cases and the total population.

```{r}
table2
```

`table3` has a column called `rate` which contains multiple values, i.e. the number of cases
and the total population. If we wanted to extract just the number of cases for each country
and year this would require some effort.

Note that the rate as a percentage could easily be computed from
`table1` and much less straightforwardly from `table2`.

```{r}
table3
```

`table4a` only contains the numbers of TB cases and `table4b` contains only the total
populations. To compute the percentage of TB cases we would need to combine these tables,
something we will look at later on in the course.

```{r}
table4a
```

```{r}
table4b
```

A dataset is a collection of values, usually either numbers (if quantitative) or text
strings (if qualitative).

Every **value** belongs to a **variable** and an **observation**.

A variable contains values that measure the same underlying attribute, e.g. height,
or temperature.

An observation contains values measured on the same unit, e.g. a person or a date.
Note that there can be multiple measurements within an observation but these should be
for multiple attributes measured on the same unit.

A tidy dataset is a data frame (or table) for which the following are true:

1. Each **variable** has its own column
2. Each **observation** has its own row
3. Each **value** has its own cell

**Question:** _which of the representations we've just been looking at is tidy?_

`table2` is tricky. If we consider the observational unit to be a country in a specific
year then `table2` is not tidy because observations are split across two rows. Also the
count variable contains counts of what are essentially different things, the number of
cases of TB and the total population. In tricky situations like this, a tell-tale sign
that your data is not in a tidy format is when you want to visualize your data and you
have to perform some kind of filtering to do so. In this case if we wanted to create a
bar plot of for the population of each country we would have to first remove the rows
corresponding to the number of TB cases.

Let's look at another untidy dataset.

```{r}
untidy_data <- read_tsv("tidyr-example.txt")
untidy_data
```

**Questions**

1. _What makes this an untidy dataset?_

2. _What other problems can you see with this dataset?_

Data in this format are quite familiar to us, but not easily interpretable by the computer.
We need to think of the dataset in terms of _variables_ and _values_.

_What are the variables in this dataset?_

* Person (John, Jane or Mary)
* Treatment (A or B)
* The effect or result of the treatment for that individual

Remember the guiding principles:

1. Each variable has its own column
2. Each observation has its own row
3. Each value has its own cell and belongs to a variable and an observation

A tidy form of this data would have 3 columns: Person, Treatment and Result.

This is an example of a fairly common scenario in which there are columns that are not variables
but instead are values of a variable.

The variable in question is `Treatment` and the two columns `treatmenta` and `treatmentb` are
values that variable can take. So a tidy form of this dataset would have a column called
`Treatment`.

### Using `gather` to create a tidy data frame

The `gather` function takes multiple "value" columns and collapses these into key-value
pairs.

Look at the help for the `gather` function to see what we need to give it.

```{r eval = FALSE}
?gather
```

We have to specify which columns to work on and we have to tell `gather` what to call the
variable for which those columns will become values - this variable is the 'key'.

The values within those 'key' columns are values but for another variable which we specify
using the 'value' argument.

In our example the key would be 'Treatment' and the value would be 'Result'.

```{r}
tidy_data <- gather(untidy_data, key = "Treatment", value = "Result", treatmenta, treatmentb)
tidy_data
```

The key and value variables can be given as strings or as symbols without quotes.

```{r}
gather(untidy_data, Treatment, Result, treatmenta, treatmentb)
```

If there were lots of treatment columns it would be quite laborious to type them all in.
Instead you can tell `gather` which columns not to include.

```{r}
gather(untidy_data, Treatment, Result, -Name)
```

You can also specify the columns as a range either using the column names

```{r}
gather(untidy_data, Treatment, Result, treatmenta:treatmentb)
```

or column indices

```{r}
gather(untidy_data, Treatment, Result, 2:3)
```

Note that `gather` is very similar to the `melt` function from the `reshape` and `reshape2` packages.

### Exercise: tidying data

See separate R markdown document.

### Other useful `tidyr` functions

There are some other useful functions within the `tidyr` package. We'll look at these using
the dataset from the exercise.

```{r}
clinical_data <- read_tsv("clinical-data.txt")
clinical_data <- gather(clinical_data, Treatment, Value, -Subject)
clinical_data
```

```{r}
ggplot(clinical_data, mapping = aes(x = Treatment, y = Value)) +
  geom_boxplot()
```

The suffix for each of the treatments, either '.1' or '.2', indicates that each drug or placebo
was measured twice. Ideally we would extract this into a new variable called 'Replicate'. We can
use the `separate` function to do so.

```{r}
clinical_replicate_data <- separate(clinical_data, Treatment, into = c("Treatment", "Replicate"))
clinical_replicate_data
```

**Question:** _how did `separate` know to split the values on the '.' character?_

Look at the help for the `separate` function and find how you would specify the delimiter. In this
case the default delimiter does the trick.

Now that we have a separate variable/column for the replicate we can create an improved box plot
with faceting.

```{r}
ggplot(clinical_replicate_data, mapping = aes(x = Replicate, y = Value)) +
  geom_boxplot() +
  facet_wrap(~Treatment)
```

The `spread` function does the opposite to `gather` and can be used to convert a tidy dataset
back to one in which we have values in column headings. Sometimes this really is necessary, e.g.
gene expression matrices where there are rows for each gene and columns for each sample, because
some analysis functions require the matrix form of the data.

```{r}
spread(clinical_data, key = Treatment, value = Value)
```


```{r}
table2
```

```{r}
spread(table2, key = type, value = count)
```

## Selecting columns

We're going to look now at some of the functions provided by the `dplyr` package, another
core component of the _tidyverse_. `dplyr` provides a set of data manipulation functions
that have a common syntax and work together in a consistent and uniform manner. These are
referred to as ***verbs*** in the `dplyr` documentation.

The most basic operation is to select a subset of columns.

First we'll load a more complex dataset with several columns (variables).

```{r}
patients <- read_tsv("patient-data.txt")
patients
```

View the patients dataset in RStudio by selecting it in the Environment tab (usually top-right) or
as follows.

```{r eval = FALSE}
View(patients)
```

You should be familiar with how to access a specific column of a data frame using the `$` operator.

```{r}
patients$Name
```

The `$` operator works just fine for single columns, but for multiple columns it quickly starts to get
cumbersome as we need to resort to using `[]` and specifying the column names in quotes.

```{r}
patients[,c("Name","Race")]
```

Non R users probably find these commands a bit obtuse.

- why the `[ ]`?
- what is `c`?
- need to remember the row and column index
- `[,...]` means display all rows

However, the `dplyr` verb `select` is a lot more intuitive.

```{r}
select(patients, Name)
```

Note that we don't need the quotation marks.

```{r}
select(patients, Name, Race)
```

We can exclude columns in the same way as we did earlier with the `gather` function.

```{r}
select(patients, -Name)
```

**Question:** _how would you do this with base R?_

```{r}
patients[,-2]
patients[,setdiff(colnames(patients), "Name")]
```

You can also select columns within a range using `:`, again as we did before with `gather`.

```{r}
select(patients, Name:Sex)
```

Looking at the help for `select` you will see that there are a number of useful functions
that can be used when selecting columns.

```{r}
select(patients, starts_with("Grade"))
```

You can combine explicit naming of columns, ranges and functions, e.g.

```{r}
select(patients, Name, Sex:Birth, -Smokes)
```

### Exercise: selecting columns

See separate R markdown document.

## Transforming and cleaning the data

We're now going to explore this dataset, taking a look at some of the variables and some
potential problems that will hinder any analysis we want to carry out.

We'll read the data in again but this time using the base R `read.delim` function instead
of the `read_tsv` because the latter has in fact done some tidying of our data and for the
purposes of what comes next it would be better if it hadn't.

```{r}
patients <- read.delim("patient-data.txt")
patients <- as_tibble(patients)
patients
```

We'll start by looking at the Sex column. Create a bar plot to show the numbers of male
and female patients in our study.

```{r}
ggplot(patients, mapping = aes(x = Sex)) +
  geom_bar()
```

There appear to be more categories than expected - why is this?

Note that the `Sex` variable has been read in as a factor. We can check the factor levels.

```{r}
levels(patients$Sex)
```

OK, so there are some extraneous spaces in the Sex column in the tabular file.

R has many functions for dealing with and manipulating text strings. However, the `stringr`
package that is part of the _tidyverse_ aims to provide a clean and uniform interface to many
common string operations.

Select the 'Packages' tab in RStudio and navigate to the `stringr` package. Take a look at the
various functions it offers and work out which one will help us remove the spaces at the beginning
and/or end of our `Sex` values.

```{r}
str_trim(patients$Sex)
```

`str_trim` has taken a vector of strings and returns another vector in which each string value
has had whitespace removed from the beginning and end of the string.

We now need a way to use this to modify the data frame. One way to do this would be:

```{r eval = FALSE}
patients$Sex <- str_trim(patients$Sex)
```

The `dplyr` function for making changes to a data frame is `mutate`.

```{r}
mutate(patients, Sex = str_trim(Sex))
```

Notice that we didn't need to specify the Sex variable as `patients$Sex`.

Usually for cleaning operations like this we would choose to overwrite the column in question
but we could instead create a new variable, e.g.

```{r}
mutate(patients, SexTrimmed = str_trim(Sex))
```

Creating a new variable is more common where we compute a new value based on values from other columns.

```{r}
mutate(table1, percent = 100 * cases / population)
```

We may want the `Sex` column to remain as a factor.

```{r}
patients <- mutate(patients, Sex = factor(str_trim(Sex)))
levels(patients$Sex)
```

```{r}
ggplot(patients, mapping = aes(x = Sex)) +
  geom_bar()
```

Let's turn our attention to another of the variables, `Height`. Create a histogram of the
heights of our patients.

```{r eval = FALSE}
ggplot(patients, mapping = aes(x = Height)) +
  geom_histogram()
```

The error is not terribly easy to understand but it does indicate that for a histogram
we need a continuous variable. What type is our `Height` variable? Why might this be?

```{r}
select(patients, Height)
```

To convert these values to numbers we need to remove the 'cm' unit that is appended
to each value.

Again, take a look at the functions available in `stringr` and see which might be able
to carry out this operation.

```{r}
str_remove(patients$Height, "cm")
```

The values are still strings though.

```{r}
as.numeric(str_remove(patients$Height, "cm"))
```

Now we can modify the dataset with `mutate`.

```{r}
patients <- mutate(patients, Height = as.numeric(str_remove(Height, "cm")))
patients
```

```{r}
ggplot(patients, mapping = aes(x = Height)) +
  geom_histogram()
```

You can modify multiple columns within a single `mutate` function call.

```{r}
patients <- read.delim("patient-data.txt")
patients <- as_tibble(patients)
patients <- mutate(patients,
                   Sex = as.factor(str_trim(Sex)),
                   Height = as.numeric(str_remove(Height, "cm")))
patients
```

### Exercise: tidying data

See separate R markdown document.

### Mutating multiple columns

If you look at the help documentation for `mutate` you will notice that there are a
set of related functions, `mutate_all`, `mutate_if` and `mutate_at`. These are very
useful functions for applying the same operation to several columns within a table
in one go.

For example, let's say we want to round each patient's weight and height to 1 decimal
place.

```{r}
patients <- read_tsv("patient-data-cleaned.txt")
mutate_at(patients, vars(Height, Weight), round, digits = 1)
```

Two things to note here.
Firstly, we had to specify the variables that needed to be modified in a `vars()` argument.
Secondly, you can add any additional arguments that the function requires at the end.

`mutate_if` allows us to run the same operation on all columns satisfying a specified criterion.
For example we could choose to round all numeric columns to 1 decimal place.

```{r}
patients <- read_tsv("patient-data-cleaned.txt")
patients <- mutate_if(patients, is.numeric, round, digits = 1)
patients
```

