---
title: "Tidying and transforming data - minimal live coding script"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---

```{r echo=FALSE}
options(tibble.print_max = 2, tibble.print_min = 2, max.print=10)
```

## Transforming and cleaning the data

```{r setup}
library(tidyverse)
```

### Read in **untidy** data

```{r}
patients <- read.delim("data/patient-data.txt")
patients <- as_tibble(patients)
patients
```

## `Sex` column

```{r echo=FALSE, fig.height=3}
ggplot(patients, mapping = aes(x = Sex)) +
  geom_bar()
```

```{r}
levels(patients$Sex)
```

### `stringr` in the 'Packages' tab

**Which function to remove spaces???**

```{r}
str_trim(patients$Sex)
```


### Modify the data frame

**base R**

```{r eval = FALSE}
patients$Sex <- str_trim(patients$Sex)
```

**dplyr**

```{r}
mutate(patients, Sex = str_trim(Sex))
```


## Creating a new variable

```{r}
mutate(patients, SexTrimmed = str_trim(Sex))
```

```{r}
mutate(table1, percent = 100 * cases / population)
```

### keep `Sex` column as a factor

```{r}
patients <- mutate(patients, Sex = factor(str_trim(Sex)))
levels(patients$Sex)
```

```{r echo=FALSE, fig.height=3}
ggplot(patients, mapping = aes(x = Sex)) +
  geom_bar()
```

## `Height`

```{r eval = FALSE}
ggplot(patients, mapping = aes(x = Height)) +
  geom_histogram()
```

```
Error: StatBin requires a continuous x variable: the x variable is discrete. Perhaps you want stat="count"?
```

### Problem is????

```{r echo=FALSE}
select(patients, Height)
```

### Which `str_`?? - **`str_remove`**

```{r}
str_remove(patients$Height, "cm")
```

### ... and to `numeric`

```{r}
as.numeric(str_remove(patients$Height, "cm"))
```

### fix the data frame

```{r}
patients <- mutate(patients, Height = as.numeric(str_remove(Height, "cm")))
patients
```

```{r echo=FALSE, fig.height=3}
ggplot(patients, mapping = aes(x = Height)) +
  geom_histogram()
```

### modify multiple columns within a single `mutate` function call.

```{r message=FALSE, warning=FALSE}
patients <- read.delim("data/patient-data.txt")
patients <- as_tibble(patients)
patients <- mutate(patients,
                   Sex = as.factor(str_trim(Sex)),
                   Height = as.numeric(str_remove(Height, "cm")))
patients
```

## Mutating multiple columns - `mutate_at` and `mutate_if`

### `mutate_at` - Round weight and height to 1 digit

```{r message=FALSE, warning=FALSE}
patients <- read_tsv("data/patient-data-cleaned.txt")
mutate_at(patients, vars(Height, Weight), round, digits = 1)
```

**or**

```{r message=FALSE, warning=FALSE}
mutate_at(patients, vars(ends_with("eight")), round, digits = 1)
```

### `mutate_if` - round all numeric columns to 1 decimal place.

```{r message=FALSE, warning=FALSE}
patients <- read_tsv("data/patient-data-cleaned.txt")
patients <- mutate_if(patients, is.numeric, round, digits = 1)
patients
```

### Also there is `mutate_all`
