---
title: "Title"
author: "Name"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---

******

Take the steps used to clean the patients dataset and calculate BMI (see below for the code)
- Re-write in the piping framework


******

```{r message = FALSE}
library(dplyr)
library(stringr)

patients <- read.delim("patient-data.txt")
patients <- tbl_df(patients)
patients_clean <- mutate(patients, Sex = factor(str_trim(Sex)))
patients_clean <- mutate(patients_clean, Height= as.numeric(str_replace_all(patients_clean$Height, pattern = "cm", "")))
patients_clean <- mutate(patients_clean, Weight = as.numeric(str_replace_all(patients_clean$Weight, "kg", "")))
patients_clean <- mutate(patients_clean, BMI = (Weight/(Height/100)^2), Overweight = BMI > 25)
patients_clean <- mutate(patients_clean, Smokes = str_replace_all(Smokes, "Yes", "TRUE"))
patients_clean <- mutate(patients_clean, Smokes = as.logical(str_replace_all(Smokes, "No", "FALSE")))
```


```{r}
## Re-write the above template using 'pipes'


```


## Exercise: filter

Use `filter` to print the following subsets of the dataset
  - Choose the Female patients from New York or New Jersey
  - Choose the overweight smokers that are still alive
  - Choose the patients who own a Pet that is not a dog
    + recall that `is.na` can check for values that are `NA`
  - (OPTIONAL)
  - Patients born in June
    + you could check out the notes on dealing with dates in the previous section for this
  - Patients with a Number > 100
    + Patient Number is the third component of the ID variable
    + e.g. `AC/AH/001` is patient Number 1; `AC/AH/017` is patient Number 17
  - Patients that entered the study on 2016-05-31
    + the column containing this data is incomplete; blank entries are assumed to be the same as the last non-blank entry above
    + the function `fill` from `tidyr` will help here

Feel free to experiment with different ways to do these


```{r}

### Your answer here
```

