---
title: "Towards Analysis Workflows in R - live coding script"
author: "Matt Eldridge and Mark Dunning"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
---

## Introduction

In this session we will introduce the concept of 'piping' to help with creating workflows
from chains of manipulations on our data. We'll also look at a couple of other useful `dplyr`
verbs.

* Introducing piping
* **filter** verb
* **arrange** verb

### Load the tidyverse

If you haven't already done so, or are working in a new session, you'll need to load
the core packages from the _tidyverse_.

```{r}
library(tidyverse)
```

## Piping

In the previous exercise we ended up making a series of manipulations to the patients
dataset.

```{r eval = FALSE}
patients <- read_tsv("patient-data.txt")
patients <- mutate(patients, Smokes = str_replace(Smokes, "No", "FALSE"))
patients <- mutate(patients, Smokes = str_replace(Smokes, "Yes", "TRUE"))
patients <- mutate(patients, Smokes = as.logical(Smokes))
patients <- mutate(patients, Height = as.numeric(str_remove(Height, pattern = "cm$")))
patients <- mutate(patients, Weight = as.numeric(str_remove(Weight, pattern = "kg$")))
patients <- mutate(patients, BMI = Weight / (Height / 100) ** 2)
patients <- mutate(patients, Overweight = BMI > 25)
```

Each statement includes an assignment to overwrite the data frame on which we are
operating. Surely this could be written in a more succinct and elegant manner.

The _tidyverse_ imports a very useful operator, `%>%` from the `magrittr` package. This
is the 'pipe' operator and works a bit like the Unix pipe operator allowing the output
from one operation to be "piped" in as the input to another operation.

Let's look at one of those cleaning operations on the patients dataset to see how piping
works.

```{r}
patients <- read_tsv("patient-data.txt")
mutate(patients, Height = as.numeric(str_remove(Height, pattern = "cm$")))
```

Instead of passing the patients data frame into the `mutate` function as it's first
argument we could use the `%>%` operator as follows.

```{r}
patients %>% mutate(Height = as.numeric(str_remove(Height, pattern = "cm$")))
```

The basic form of a piped operation is:

**`x %>% f(y)` is equivalent to `f(x, y)`**

Piping becomes really useful when there are a number of steps involved in transforming
a dataset.

```{r}
patients <- read.delim("patient-data.txt") %>%
  as_tibble %>%
  mutate(Sex = as_factor(str_trim(Sex))) %>%
  mutate(Height = as.numeric(str_remove(Height, pattern = "cm$")))
```

The usual way of developing a workflow is to build it up one step at a time, testing the
output produced at each stage.

### Exercise: re-writing a workflow using pipes

See separate markdown document.

## Filtering rows

The **`filter`** verb allows you to choose rows from a data frame that match some specified
criteria. The criteria are based on values of variables and can make use of comparison
operators such as `==`, `>`, `<` and `!=`.

For example to filter the patients dataset so it only contains males.

```{r}
patients <- read_tsv("patient-data-cleaned.txt")
filter(patients, Sex == "Male")
```

The equivalent in base R is much less intuitive.

```{r}
patients[patients$Sex == "Male",]
```

We can also use the `!=` operator.

```{r}
filter(patients, Sex != "Female")
```

We can filter for a set of values using the `%in%` operator.

```{r}
filter(patients, State %in% c("Florida", "Georgia", "Illinois"))
```

Partial matches can be made using the `str_detect` function from the `stringr` package.
Note this is similar to the `grepl` function in base R.

For example, let's select all the patients whose name begins with a 'B'.

```{r}
filter(patients, str_detect(Name, "^B"))
```

Note that the `str_detect` function returns a logical vector - this is important since
the criterion for filtering must evaluate to `TRUE` or `FALSE`.

Also note that the second argument to `str_detect` is a regular expression. An alternative
function from `stringr` we could have used in this case is `str_starts`; with this we no
longer need to '^' symbol in our regular expression.

```{r}
filter(patients, str_starts(Name, "B"))

```

We can filter on logical variables straightforwardly.

```{r}
filter(patients, !Died)
```

Also we can add extra conditions, separating them with a `,`.

```{r}
filter(patients, Sex == "Male", Died)
```

In this example all males who have died are selected. The `,` is the equivalent of using
the Boolean operator `&`.

```{r}
filter(patients, Sex == "Male" & Died)
```

The equivalent in regular R is more verbose and less easy to read.

```{r}
patients[patients$Sex == "Male" & patients$Died,]
```

We can use the `|` Boolean operator to select patients above a given weight or BMI.

```{r}
filter(patients, Weight > 90 | BMI > 28)
```

Mixing both Boolean operators and `,` is also possible.

```{r}
filter(patients, Weight > 90 | BMI > 28, Sex == "Female")
```

### Exercise: filtering rows

See separate markdown document.

## Sorting rows

Another `dplyr` verb that works on rows in a table is **`arrange`**. This is used to sort
rows in a dataset based on one or more variables.

For example, let's say we want to sort our patients by height.

```{r}
arrange(patients, Height)
```

This has arranged the rows in order of ascending height. What if we wanted descending order
of height?

```{r}
arrange(patients, desc(Height))
```

We can use sort using multiple variables, e.g. first by Grade in descending order, then by Sex and then Smokes.

```{r}
arrange(patients, desc(Grade), Sex, Smokes)
```

Sorting is commonly used in workflows usually as one of the last steps before presentation or
writing out the resulting table to a file.

The following concise and easy-to-read workflow includes steps that use all of the `dplyr` verbs we have covered so
far.

```{r}
candidates <- patients %>%
  filter(!Died) %>%
  select(ID, Name, Sex, Smokes, Height, Weight) %>%
  mutate(BMI = Weight / (Height / 100) ** 2) %>%
  mutate(Overweight = BMI > 25) %>%
  arrange(BMI)
candidates
```


