---
title: "Summarizing and Grouping - minimal live coding script"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---

```{r echo=FALSE}
options(tibble.print_max = 2, tibble.print_min = 2, max.print=10)
```

## Recap

### Tidy data

A tidy dataset is a data frame (or table) for which the following are true:

1. Each **variable** has its own column
2. Each **observation** has its own row
3. Each **value** has its own cell

### dplyr basics

There are 5 key dplyr functions with which the vast majority of data manipulations can be achieved.

* `select` -- select variables (columns) of interest
* `filter` -- pick observations (rows) based on their values
* `arrange` -- reorder the observations based on one or more variables
* `mutate` -- modify existing or create new variables by applying functions to existing variables
* `summarize` -- collapse many values to a single summary

We've covered the first four and will focus now on summarizing.

## Summarizing

```{r}
library(tidyverse)
```

```{r}
patients <- read_tsv("data/patient-data-cleaned.txt")
patients
```

How do you compute the mean height of patients in the dataset?

```{r}
mean(patients$Height)
```

This has collapsed a vector of height values to a summary, i.e. the average value.
The `dplyr` function or 'verb' for doing this is `summarize` (or `summarise`).

```{r}
summarize(patients, mean(Height))
```

This is not particularly useful on it's own but you can compute multiple summarized values at the same time.

```{r}
summarize(patients, mean(Height), sd(Height))
```

We can rename the column headings.

```{r}
summarize(patients, mean = mean(Height), standard_deviation = sd(Height))
```

We might also want to compute summaries on multiple columns. Here we can use the `summarize_at`
or `summarize_if` functions that work in a very similar way to `mutate_at` and `mutate_if` that
we've come across earlier

```{r}
summarize_at(patients, vars(Height, Weight), mean)
```

We can also use the same helper functions we've used before with `select`.

```{r}
summarise_at(patients, vars(matches('[HW]eight')), mean)
```

```{r}
summarize_if(patients, is.numeric, mean)
```





Why does the Grade column give a result of `NA`?

What do we need to do to fix this?

```{r}
summarize_if(patients, is.numeric, mean, na.rm = TRUE)
```

We can compute multiple summarized values on multiple columns.

```{r}
summarize_at(patients, vars(Height, Weight), list(mean, sd))
```

And again but with more ascetically pleasing column headings.

```{r}
summarize_at(patients, vars(Height, Weight), list(mean = mean, sd = sd))
```

## Grouping

`summarize` only really becomes useful when working with subsets of the data, or groups.
Here we use the `group_by` function and it really helps if we do so in the context of a
workflow using the pipe mechanism (`%>%`).


For example, let's compute the average height of males and females separately.

```{r}
patients %>%
  summarize(mean(Height))
```


```{r}
patients %>%
  group_by(Sex) %>%
  summarize(mean(Height))
```

```{r}
patients %>%
  group_by(Sex) %>%
  summarize(`Average height` = mean(Height))
```


```{r}
patients %>%
  group_by(Sex) %>%
  summarize_at(vars(Height, Weight), mean)
```

```{r}
patients %>%
  group_by(Sex) %>%
  summarize_at(vars(Height, Weight), list(mean = mean, stddev = sd))
```

### Grouping by multiple variables

```{r}
patients %>%
  group_by(Sex, State) %>%
  summarize(
    Mean_height = mean(Height),
    Stddev_height = sd(Height)
  )
```

It would also be helpful to know how many observations of each group there are.

```{r}
patients %>%
  group_by(Sex, State) %>%
  summarize(
    Mean_height = mean(Height),
    Stddev_height = sd(Height),
    N = n()
  )
```

And finally we don't really need all those decimal places.

```{r}
patients %>%
  group_by(Sex, State) %>%
  summarize(
    Mean_height = mean(Height),
    Stddev_height = sd(Height),
    N = n()
  ) %>%
  mutate_at(vars(Mean_height, Stddev_height), round, digits = 1)
```

### Exercise

Compute the median weight and BMI and the interquartile ranges for smokers and non-smokers.

```{r}
patients %>%
  group_by(Smokes) %>%
  summarize_at(vars(Weight, BMI), list(median = median, IQR = IQR))
```

Compute the numbers of patients with a given Grade separately for each State.

```{r}
patients %>%
  group_by(State, Grade) %>%
  summarize(N = n())
```

There is a more concise way of doing this using the `count` function.

```{r}
count(patients, State, Grade)
```

### Ungrouping

It is sometimes necessary to use `ungroup` when summarizing based on grouping by multiple variables
particularly when the results are going to be further manipulated.

```{r}
summary <- patients %>%
  group_by(Sex, State) %>%
  summarize(Height = mean(Height), Weight = mean(Weight))
summary
```

```{r}
summary %>%
  summarize_if(is.numeric, median)
```

What happened here?

It seems that the `summary` object is still grouped in some way.

```{r}
groups(summary)
```

```{r}
class(summary)
```

Let's repeat the original operation but this time explicitly remove the grouping after
sumarizing.

```{r}
summary <- patients %>%
  group_by(Sex, State) %>%
  summarize(Height = mean(Height), Weight = mean(Weight)) %>%
  ungroup()
summary
```

```{r}
groups(summary)
```

```{r}
class(summary)
```

```{r}
summary %>%
  summarize_if(is.numeric, median)
```

Why do `group_by` and `summarize` behave like this? There are some scenarios where this is helpful.

```{r}
diabetes <- read_tsv("data/diabetes.txt")
diabetes
```

```{r}
diabetes <- diabetes %>%
  separate(Date, into = c("year", "month", "day"), sep = "-")
diabetes
```

```{r}
daily <- group_by(diabetes, year, month, day)
groups(daily)
```

```{r}
per_day <- summarize(daily, measurements = n())
groups(per_day)
per_day
```

```{r}
per_month <- summarize(per_day, measurements = sum(measurements))
groups(per_month)
per_month
```

```{r}
per_year <- summarize(per_month, measurements = sum(measurements))
groups(per_year)
per_year
```

Each time we peel away one layer of grouping.

**Tip**: always use `ungroup` after grouping on multiple variables
