---
title: "Programming: functions and loops -- minimalscript"
author: "Ashley Sawle"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
---

```{r include = FALSE}
library(tidyverse)
options(tibble.print_max = 3, tibble.print_min = 3)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

### Read the patients data into R environment

```{r setup}

patients <- read_tsv("data/patient-data-cleaned.txt") %>%  
    select(Height, Weight, -BMI)
```

## Writing our own functions

### Basic syntax to generate a function

Lets look at a tidyverse function definition

```{r topn, eval=FALSE}
top_n
```

* The command to create a function is `function`
* This is followed by the arguments of the function in brackets
* Finally the code that the function will execute is between curly brackets

So the general syntax is:

```
Function_Name <- function(arg1, arg2, arg3, ...){
    Code_Line_1
    Code_Line_2
    ...
}
```

Let's create a simple function. This one has no arguments.

```{r firstFunc, message=TRUE}
my_first_function  <- function(){
    message("Hello World!")
}

my_first_function()
```

Let's make it a bit more useful by adding an argument

```{r firstFunc2, message=TRUE}
sayHello <- function(name){
    message("Hello ", name, "!")
}

sayHello("Ash")
sayHello("Matt")
```

### Functions with output

So far our function just printed a message to the console, but most functions return
a value.

In R a function can only return a single object

Let's create a function to calculate BMI: weight (kg) / height^2^ (m) --> kg/m^2^

```{r BMIfunc}
bmi <- function(height, weight){
    weight / (height ^ 2)
}
bmi(1.75, 78)
```

Let's try that with our patients table.
```{r testBMI}
patients  %>%  
    mutate(BMI=bmi(Height, Weight))
```

This isn't right because the height should be in metres. Let's add some
additional arguments to define the units and then some extra code to adjust the
values depending on the units provided.

### Defaults

```{r BMIfunc2}
bmi <- function(height, weight, height_units="m", weight_units="kg"){
    if(height_units == "cm"){ height <- height / 100 }
    if(weight_units == "g"){ weight <- weight / 1000 }
    weight / (height ^ 2)
}

bmi(1.75, 78)
patients  %>%  
    mutate(BMI=bmi(Height, Weight, height_units = "cm"))
```

### Returned object

There is a function `return` but we don't usually need it.

```{r}
bmi <- function(height, weight, height_units="m", weight_units="kg"){
    if(height_units == "cm"){ height <- height / 100 }
    if(weight_units == "g"){ weight <- weight / 1000 }
    bmi <- weight / (height ^ 2)
    return(bmi)
}
bmi(165, 67)
```

### Generating a plot with a function


```{r}
diabetes <- read_tsv("data/diabetes.txt")
```

```{r single_patient_plot, fig.height=2}
diabetes %>%  
    filter(ID == "AC/AH/001") %>%  
    ggplot(aes(x = Date, y = Glucose)) +
        geom_line(colour = "blue") +
        labs(title = "Patient: AC/AH/001")
```

How can we turn this into a function? Step by step...

```{r single_patient_plot2, fig.height=2}
patientID <- "AC/AH/001"
plotTitle <- str_c("Patient: ", patientID)

diabetes %>%  
    filter(ID == patientID) %>%  
    ggplot(aes(x = Date, y = Glucose)) +
        geom_line(colour = "blue") +
        labs(title = plotTitle)
```

```{r plotAsAFunction, fig.height=2}
plotGlucose <- function(patientID, dat=diabetes, myColour="blue"){
    plotTitle <- str_c("Patient: ", patientID)
    dat %>%  
        filter(ID == patientID) %>%  
        ggplot(aes(x = Date, y = Glucose)) +
            geom_line(colour = myColour) +
            labs(title = plotTitle)
}

plotGlucose("AC/AH/104")
plot1 <- plotGlucose("AC/AH/017", myColour="purple")
plot1
```

## Doing repetitive tasks with loops

R works nicely with vectors
```{r}
x <- c(1, 2, 3, 4, 5)
log2(x)
```

but not always the behaviour we want

```{r}
sum(x, 5)
```

so we could do...

```{r eval=FALSE}
sum(1, 5)
sum(2, 5)
sum(3, 5)
sum(4, 5)
sum(5, 5)
```

If you are repeating code like this, you should use loop.

### Basic syntax

The basic syntax of a `for` loop is:

```
for(a_variable in a_vector){
    some code operating on a_variable
}
```

Let's look at a simple example:

```{r}
for(a_number in c(11, 21, 16, 4, 500)){
    sum(a_number, 5)
}
```

In this case the code has been run 5 times...

More commonly we would define the vector first:

```{r}
some_data <- c(11, 21, 16, 4, 500)
for(a_number in some_data){
    sum(a_number, 5)
}
```

often we will use a sequence

```{r message=TRUE}
some_data <- c(11, 21, 16, 4, 500)
for(a_number in seq(1, 5)){
    new_number <- sum(some_data[a_number], 5)
    message("Result number ", a_number, " is ", new_number)
}
```

A better way is to use `seq_along`

```{r message=TRUE}
some_data <- c(11, 21, 16, 4, 500)
for(a_number in seq_along(some_data)){
    new_number <- sum(some_data[a_number], 5)
    message("Result number ", a_number, " is ", new_number)
}
```

### A more realistic example

Let's say that with our patients data we want to export a separate table for
each state, containing just the men, with ID, Name, Smokes, BMI and their their
average BP from the diabetes table.

```{r eval=FALSE}
states <- unique(patients$State)

avg_bp  <- diabetes %>%  
    group_by(ID) %>%  
    summarise(AvgBP=mean(BP)) %>%  
    ungroup()

for(state in states){
    outputFile <- str_c("data/", state, "_male_patient_data.tsv") %>%  
        str_replace(" ", "_")
    patients  %>%  
        filter(State == state & Sex == "Male") %>%  
        left_join(avg_bp, by = "ID") %>%  
        select(ID, Name, Smokes, BMI, AvgBP) %>%  
        write_tsv(outputFile)
}
```

### Emma's problem - A loop with plotting

There are many ways to solve this problem; this is just one of them.

```{r eval=FALSE}
all_patients <- unique(diabetes$ID)
len <- length(all_patients)
start_index <- seq(1, len, by = 16)

plotGlucose <- function(dat){
    ggplot(dat, aes(x = Date, y = Glucose)) +
        geom_line(colour = "red") +
        geom_point(shape = 21, fill = "red") +
        facet_wrap(~ID, ncol=4)
}

pdf("data/My_plots.pdf")
for(stt in start_index){
    end <- min(stt + 15, 100)
    these_patients <- all_patients[seq(stt, end)]
    diabetes %>%  
        filter(ID %in% these_patients) %>%  
        plotGlucose() %>%  
        print()
}
dev.off()
```

The last page isn't ideal perhaps. We could solve this with some more advanced
plotting using `gridExtra` to create the grid of plots from a `list` of
individual plots. We will also need a nested loop - a loop within a loop.

```{r eval=FALSE}
library(gridExtra)

pdf("data/My_plots.pdf")
for(stt in start_index){
    end <- min(stt + 15, 100)
    these_patients <- all_patients[seq(stt, end)]
    plotList <- list() 
    for(i in seq_along(these_patients)){
        glPlot <- diabetes %>%  
            filter(ID %in% these_patients[i]) %>%  
            plotGlucose()
        plotList[[i]] <- glPlot
    }
    grid.arrange(grobs = plotList, ncol = 4, nrow = 4)
}
dev.off()

```


























