---
title: "Exercise Set 1 — Geoms and Aesthetics"
author: "Mark Dunning and Matt Eldridge"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---


```{r include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, eval = TRUE)
```


## Exercise 1

These first few exercises will run through some of the simple principles of creating a ggplot2 object, assigning aesthetics mappings and geoms.


1. Read in the cleaned patients dataset as we saw in ggplot2 course earlier.

```{r}
library(tidyverse)
patients <- read_tsv("patient-data-cleaned.txt")
```

### Scatterplots

2. Using the patient dataset generate a scatter plot of BMI versus Weight.

```{r exercise1}
ggplot(data = patients, mapping = aes(x = BMI, y = Weight)) +
  geom_point()
```

3. Extending the plot from exercise 2, add a colour scale to the scatterplot based on the Height variable.

```{r exercise2}
ggplot(data = patients, mapping = aes(x = BMI, y = Weight, colour = Height)) +
  geom_point()
```

4. Following from exercise 3, split the BMI vs Weight plot into a grid of plots separated by Smoking status and Sex.

```{r exercise3}
ggplot(data = patients, mapping = aes(x = BMI, y = Weight, colour = Height)) +
  geom_point() +
  facet_grid(Sex ~ Smokes)
```

5. Using an additional geom, add an extra layer of a fit line to the solution from exercise 3.

```{r exercise4}
ggplot(data = patients, mapping = aes(x = BMI, y = Weight, colour = Height)) +
  geom_point() +
  geom_smooth()
```

6. Does the fit in question 5 look good? Look at the description for ?geom_smooth() and adjust the method to fit a straight line without standard error bounds.

```{r exercise5}
ggplot(data = patients, mapping = aes(x = BMI, y = Weight, colour = Height)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

### Boxplots and Violin plots

7. Generate a boxplot of BMIs comparing smokers and non-smokers.

```{r exercise6}
ggplot(data = patients, mapping = aes(x = Smokes, y = BMI)) +
  geom_boxplot()
```


8. Following from the boxplot comparing smokers and non-smokers in exercise 7, colour boxplot edges by Sex.

```{r exercise7}
ggplot(data = patients, mapping = aes(x = Smokes, y = BMI, colour = Sex)) +
  geom_boxplot()
```

9. Now reproduce the boxplots in exercise 8 (grouped by smoker, coloured by sex) but now include a separate facet for people of different age (using Age column).

```{r exercise8}
ggplot(data = patients, mapping = aes(x = Smokes, y = BMI, colour = Sex)) +
  geom_boxplot() +
  facet_wrap(~ Age)
```

10. Produce a similar boxplot of BMIs but this time group data by Sex, colour by Age and facet by Smoking status.

_Hint: is Age currently the right type of variable for grouping in a box plot?_

```{r exercise9}
patients$Age <- factor(patients$Age)
ggplot(data = patients, mapping = aes(x = Sex, y = BMI, colour = Age)) +
  geom_boxplot() +
  facet_wrap(~ Smokes)
```

11. Regenerate the solution to exercise 10 but this time using a violin plot.

```{r exercise10}
ggplot(data = patients, mapping = aes(x = Sex, y = BMI, colour = Age)) +
  geom_violin() +
  facet_wrap(~ Smokes)
```


### Histogram and Density plots

12. Generate a histogram of BMIs with each bar coloured blue, choosing a suitable bin width.

```{r exercise11}
ggplot(data = patients, mapping = aes(x = BMI)) +
  geom_histogram(fill = "blue", binwidth = 0.5)
```

13. Generate density plots of BMIs coloured by Sex.

_Hint: alpha can be used to control transparancy._

```{r exercise12}
ggplot(data = patients, mapping = aes(x = BMI)) +
  geom_density(aes(fill = Sex), alpha = 0.5)
```

14. Generate a separate density plot of BMI coloured by sex for each Grade.
```{r exercise13}
ggplot(data = patients, mapping = aes(x = BMI)) +
  geom_density(aes(fill = Sex), alpha = 0.5) +
  facet_wrap(~ Grade)
```


<br>


