---
title: "Session 2 exercises"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Part I -- ggplot - Facets

The first part requires you to create plots with faceting. Each of the plots you are asked to create is shown below so that you can compare the end result with your own.

1. Read in the cleaned patients dataset, `patient-data-cleaned.txt`.

```{r}
library(tidyverse)
patients <- read_tsv("data/patient-data-cleaned.txt")
patients
```

1. Using the patient dataset generate a scatter plot of BMI versus Weight, add a
colour scale to the scatterplot based on the Height variable, and split the plot 
into a grid of plots separated by Smoking status and Sex.

```{r}
ggplot(data = patients, mapping = aes(x = BMI, y = Weight, colour = Height)) +
  geom_point() +
  facet_grid(Sex ~ Smokes)
```

2. Generate a boxplot of BMIs comparing smokers and non-smokers, colour boxplot
by Sex, and include a separate facet for people of different age.

```{r}
ggplot(data = patients, mapping = aes(x = Smokes, y = BMI, fill = Sex)) +
  geom_boxplot() +
  facet_wrap(~ Age)
```

3. Produce a similar boxplot of BMIs but this time group data by Sex, colour by Age and facet by Smoking status.

```{r}
patients$Age <- factor(patients$Age)
ggplot(data = patients, mapping = aes(x = Sex, y = BMI, fill = Age)) +
  geom_boxplot() +
  facet_wrap(~ Smokes)
```

4. Regenerate the solution to exercise 3 but this time using a violin plot.

```{r}
ggplot(data = patients, mapping = aes(x = Sex, y = BMI, fill = Age)) +
  geom_violin() +
  facet_wrap(~ Smokes)
```

5. Generate density plots of BMIs coloured by Sex, and split the plot by Grade.

```{r}
ggplot(data = patients, mapping = aes(x = BMI)) +
  geom_density(aes(fill = Sex), alpha = 0.5) +
  facet_wrap(~ Grade)
```

## Part II -- tidyr - Tidying a dataset

1. Read in the simulated clinical dataset, `clinical-data.txt`.

```{r}
library(tidyverse)
clinical_data <- read_tsv("data/clinical-data.txt")
clinical_data
```

2. What are the variables in this data set?

3. Transform the data into a tidy form using the `gather` function from the `tidyr` package.

```{r}
clinical_data <- gather(clinical_data, key = "Treatment", value = "Value", -Subject)
```

4. Display the range of values for each drug and placebo treatment as a box plot

```{r}
ggplot(clinical_data, mapping = aes(x = Treatment, y = Value)) +
  geom_boxplot()
```
