---
title: "Session 5 exercise: building a gene symbol lookup"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---

We commonly work with gene symbols but these are not a very precise way of referring
to specific genes, with the same symbol used in some cases as an alias for more than
one gene.

There are lots of ways of mapping gene symbols to gene entries in popular databases
or resouces, such as Ensembl or RefSeq, but in this exercise we will use a resource
provided by the HUGO Gene Nomenclature Committee.

The aim is to create a lookup that gives Ensembl and RefSeq accession ids for
gene symbols.

1. Download the HGNC complete gene set

In a web browser, search for and navigate to the HGNC website. Go to the 'Statistics and
download files' page from the 'Downloads' drop-down menu. Towards the bottom of the page
is a section called 'Complete dataset download links'. Download the Complete HGNC dataset
as a tab-separated text file ('TXT' button). You may need to right click and choose
'Save link as' depending on what brower you're using. The file is called `hgnc_complete_set.txt`.

2. Load the `hgnc_complete_set.txt` file into a data frame

```{r}
library(tidyverse)
genes <- read_tsv("hgnc_complete_set.txt")
```

3. Select the columns of interest

These are the HGNC id, the RefSeq and Ensembl accession numbers, and 3 columns containing
gene symbols - one for the primary symbol recognized by HGNC, another for symbols that were used
as the primary symbol in previous releases of the resource, and one final column containing
aliases.

It will be useful at this point to rename the 3 gene symbol columns to 'primary', 'previous'
and 'alias'

```{r}
genes <- genes %>%
  select(
    hgnc_id,
    refseq_accession,
    ensembl_gene_id,
    primary = symbol,
    previous = prev_symbol,
    alias = alias_symbol
  )
```

4. Gather the 3 gene symbol columns into a single column

Use the `gather` function to create 'symbol' and 'class' columns.

It would be very helpful here to retain the primary HGNC gene symbol column
so apply a `mutate` to copy this column to a new column called 'hgnc_symbol'
before calling `gather`.

```{r}
genes <- genes %>%
  mutate(hgnc_symbol = primary) %>%
  gather(key = class, value = symbol, primary, previous, alias, na.rm = TRUE)
```

5. Filter for missing values.

Genes with no previously used symbol or aliases will have rows with `NA`
values in the symbol column. Remove these.

```{r}
genes <- genes %>%
  filter(!is.na(symbol))
```

Note that this step is not necessary if `na.rm = TRUE` used in `gather` previously.

6. Split multiple gene symbols into separate rows.

Some genes have multiple gene symbol aliases that are separated by the '|' character.
These need to be split into separate rows. Look up the list of functions provided by
the tidyr package using the Help tab in RStudio and find the function you need to do
this.

The separator you will need to provide is a regular expression and because the '|'
character means something special within regular expression you will need to tell it
that you really mean the '|' character by prefixing it with "\\", i.e. `sep = "\\|"`).

```{r}
genes <- genes %>%
  separate_rows(symbol, sep = "\\|")
```

7. Assign an order of precedence base on the class of the symbol

Some gene symbols map to multiple genes. A primary gene symbol should take
precedence over a previously used symbol and an alias should be lower priority
still. To enable us to prioritize gene symbol mappings based on this ordering
we need to turn our 'class' column into a factor with the levels defining the
order or precedence.

```{r}
genes <- genes %>%
  mutate(class = factor(class, levels = c("primary", "previous", "alias")))
```

8. Assign a ranking to the mapping for each gene symbol based on the order of precedence

Each gene symbol needs to be treated separately, i.e. as a group. Within each group
the rows need to be sorted by the order of precedence of the type of symbol and then
a rank or priority assigned where 1 would be our favoured mapping, 2 our next preferred,
etc.

**Hint:** use the `row_number` function

```{r}
lookup <- genes %>%
  group_by(symbol) %>%
  arrange(class) %>%
  mutate(ranking = row_number()) %>%
  ungroup() %>%
  select(symbol, ranking, class, everything()) %>%
  arrange(symbol, ranking)
```

Check if you need to remove grouping in your data frame after adding the ranking.

9. Use the gene symbol lookup table

We now have our gene symbol lookup table and can use it. See which genes the following
symbols map to.

* KRAS
* MTCP1
* CASP
* HSR1
* your favourite gene

```{r}
lookup %>%
  filter(symbol %in% c("KRAS", "MTCP1", "CASP", "HSR1", "TP53"))
```

Note that using the `row_number` function means that matches for symbols with the same class
are still given different rankings.

```{r}
lookup %>%
  filter(symbol == "CYP2A")
```

We may instead want to have ties for those matches with
the same class and can use the `min_rank` function for this.

```{r}
lookup <- genes %>%
  group_by(symbol) %>%
  mutate(ranking = min_rank(class)) %>%
  ungroup() %>%
  select(symbol, ranking, class, everything()) %>%
  arrange(symbol, ranking)

lookup %>%
  filter(symbol == "CYP2A")
```

10. Using just the preferred mappings, compute the number of symbols in each class and the percentage of those that map to a HGNC gene with an Ensembl cross-reference

```{r}
lookup %>%
  filter(ranking == 1) %>%
  group_by(class) %>%
  summarize(n = n(), percent = 100 * mean(!is.na(ensembl_gene_id)))
```

