---
title: "Summarising, Grouping and Joining Exercise"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
---

### Summarising

The first part of the exercise uses the patients dataset we've been using in previous sections of the course. After reading this into R, answer the following questions using the `summarise`, `summarise_each` and `mutate_each` functions.

```{r message=FALSE}
library(dplyr)
patients <- read.delim("patient-data-cleaned.txt", stringsAsFactors = FALSE) %>% tbl_df
patients %>% head(5)
```

******

Compute the mean age, height and weight of patients in the patients dataset

- Modify the output by adding a step to round to 2 decimal places

******

```{r}
patients %>%
  summarize_each(funs(mean), Age, Height, Weight) %>%
  mutate_each(funs(round(., digits = 2)))
```

******

See what happens if you try to compute the mean of a logical (boolean) variable

- What proportion of our patient cohort has died?

******

```{r}
patients %>% summarize(mean(Died))
```

******

### Grouping

The following questions require grouping of patients based on one or more attributes using the `group_by` function.

Compare the average height of males and females in this patient cohort.

Are smokers heavier or lighter on average than non-smokers in this dataset?

******

```{r}
patients %>%
  group_by(Sex) %>%
  summarize(`Average height` = mean(Height))

patients %>%
  group_by(Smokes) %>%
  summarize(`Average weight` = mean(Weight))

patients %>%
  group_by(Sex, Smokes) %>%
  summarize(`Average weight` = mean(Weight))
```

******

### Joining

The patients are all part of a diabetes study and have had their blood glucose concentration and diastolic blood pressure measured on several dates.

This part of the exercise combines grouping, summarisation and joining operations to connect the diabetes study data to the patients table we've already been working with.

```{r}
diabetes <- read.delim("diabetes.txt", stringsAsFactors = FALSE) %>% tbl_df
diabetes %>% head(5)
```

******

Find the number of visits made by each patient to the hospital to have blood glucose and pressure measurements taken.

*[hint: use the `n` function from `dyplr`, type `?n` for help]*

Find the date of the last visit made by each patient.

*[hint: use `arrange` to sort by date and the `last` function to select the last one]*

Using the ID column to connect information about patients, add these values as additional columns to the patients table.

- Can you do this as a single sequence of dplyr commands using the `%>%` symbol?

******

```{r}
diabetes %>%
  group_by(ID) %>%
  arrange(Date) %>%
  summarize(Visits = n(), LastVisit = last(Date)) %>%
  right_join(patients, by = "ID")
```
