---
title: "Patient Data Exploration"
author: "Mark Dunning"
date: "21 April 2016"
output: html_document
---

******

Take the steps used to clean the patients dataset and calculate BMI (see template for the code)

- Re-write in the piping framework
- Add a step to print just the ID, Name, Date of Birth, Smokes and Overweight columns

******

```{r}
library(dplyr)
library(stringr)
patients_clean <- read.delim("patient-data.txt") %>% 
  tbl_df %>% 
  mutate(Sex = factor(str_trim(Sex))) %>% 
  mutate(Height= as.numeric(str_replace_all(Height,pattern = "cm",""))) %>% 
  mutate(Weight = as.numeric(str_replace_all(Weight,"kg",""))) %>% 
  mutate(BMI = (Weight/(Height/100)^2), Overweight = BMI > 25) %>% 
  mutate(Smokes = str_replace_all(Smokes, "Yes", TRUE)) %>% 
  mutate(Smokes = as.logical(str_replace_all(Smokes, "No", FALSE))) %>% 
  select(ID, Name, Birth,BMI,Smokes,Overweight,Died)
patients_clean
```

******

Modify the workflow 

- to select the candidates (overweight smokers that are still alive)
- write the result to a file

******

```{r}

  candidates <- filter(patients_clean, Smokes & Overweight & !Died) %>% 
  write.table("candidates.txt")
```


```{r}
read.delim("patient-data.txt") %>% 
  tbl_df %>% 
  mutate(Sex = factor(str_trim(Sex))) %>% 
  mutate(Height= as.numeric(str_replace_all(Height,pattern = "cm",""))) %>% 
  mutate(Weight = as.numeric(str_replace_all(Weight,"kg",""))) %>% 
  mutate(BMI = (Weight/(Height/100)^2), Overweight = BMI > 25) %>% 
  mutate(Smokes = str_replace_all(Smokes, "Yes", TRUE)) %>% 
  mutate(Smokes = as.logical(str_replace_all(Smokes, "No", FALSE))) %>% 
  select(ID, Name, Birth,BMI,Smokes,Overweight,Died) %>% 
  filter(Smokes & Overweight & !Died) %>% 
  write.table("candidates.txt")
```


******

Modify the workflow to 

- Add the Age variable to the final table
    + do you notice anything unusual about the values for `Age`?
- Arrange by age
- Select the top 5 oldest patients

******

```{r}
read.delim("patient-data.txt") %>% 
  tbl_df %>% 
  mutate(Sex = factor(str_trim(Sex))) %>% 
  mutate(Height= as.numeric(str_replace_all(Height,pattern = "cm",""))) %>% 
  mutate(Weight = as.numeric(str_replace_all(Weight,"kg",""))) %>% 
  mutate(BMI = (Weight/(Height/100)^2), Overweight = BMI > 25) %>% 
  mutate(Smokes = str_replace_all(Smokes, "Yes", TRUE)) %>% 
  mutate(Smokes = as.logical(str_replace_all(Smokes, "No", FALSE))) %>% 
  select(ID, Name, Birth,BMI,Smokes,Overweight,Died,Age) %>% 
  filter(Smokes & Overweight & !Died) %>% 
  filter(Age > 0) %>% 
  top_n(5,Age) %>% 
  arrange(desc(Age))
```

