---
title: "R Notebook"
output: html_notebook
---

## ggplot2 further practice
We will be using a Breast cancer biopsy data-set from the Wisconsin Hospitals.
This is available from a package called MASS so you will need to install & load it along with the tidyverse package.
```{r}
install.packages("tidyverse")
install.packages("MASS")
library(tidyverse)
library(MASS)
```
The MASS package loads functions and dataframes. The dataframe of interest is called biopsy.
Use the R help function to find out more about this data frame and then you should examine a few rows of the data and observe the data types for each column.
```{r}
?biopsy

biopsy
```
There are 699 rows of data. We can start by visualising the relative numbers of benign & malignant tumours using a bar-plot (try colouring the bars).
```{r}
ggplot(data=biopsy) + geom_bar(mapping = aes(x = class, fill = class))
```
Let's now produce some histograms to explore the distribution of some of our datasets variables.
Try clump thickness & then normal nuclei. Explore effects of various bin numbers and binwidths
(use ?ggplot to get some hints)
```{r}
ggplot(data=biopsy) + geom_histogram(mapping=aes(x=V1))
ggplot(data=biopsy) + geom_histogram(mapping=aes(x=V8))
```
How might we split (or facet - hint,hint) our data into two plots one for benign and one for malignant? (You may need to use the R help function).
```{r}
ggplot(data=biopsy) + geom_histogram(mapping=aes(x=V1)) + facet_grid(cols=vars(class))
```
Let's now explore the data variability of clump thickness in benign and malignant cases using a box-plot. Feel free to plot the other variables e.g. uniformity.

```{r}
ggplot(data=biopsy, mapping= aes(x=class, y=V1)) + geom_boxplot() + geom_point()
```
Try and improve this plot using jitter to see how much overlapping is present.
```{r}
ggplot(data=biopsy, mapping= aes(x=class, y=V1)) + geom_boxplot() + geom_jitter() 
```
OK - so that's a bit messy maybe we'd be better off with a violin plot?
```{r}
ggplot(data=biopsy, mapping= aes(x=class, y=V1)) + geom_violin() + geom_jitter()
```


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

