---
title: "Further dplyr practice"
output:
  html_document:
    df_print: paged
  code_folding: hide
  html_notebook: null
---

This is a set of extra exercises to further assess if you are getting familiar with **dplyr**.   
Please try and do the tasks without looking at the provided code and note that there may be    
other ways to achieve the task than the one provided. Use the code button to the left and    select 'hide all code'.    

## Setting up

We will be using an Ovarian Cancer dataset that is included in the **vcd** package. You will need to install this (along with **tidyverse** if you have not already done so) and then load (using **library**) both packages.

```{r}
options(repos = c("CRAN" = "http://cran.ma.imperial.ac.uk"))
install.packages("vcd")
library(vcd)
library(tidyverse)
```

Now we need to load the dataset as an R dataframe using the **data()** command.
Take a look at the dataframes data. Use **as_tibble()** to view it as a tibble and
note the difference. Finally, use the **as_tibble()** to store the dataframe as a tibble
in a new variable for later usage.
```{r}
data("OvaryCancer")

?OvaryCancer

OvaryCancer

as_tibble(OvaryCancer)

ovdata <- as_tibble(OvaryCancer)
```
## Putting dplyr to work on the OvaryCancer dataset   
It might be useful to have a % of the total number of patients for our categories.  
We can use **mutate()** after using **select()** & **sum()** to determine the total number of   
patients in the study.    
Devise the code needed.    
```{r}
pat_total <- sum(select(ovdata, Freq))
pat_total
ovdata %>%
  mutate(pc_surv = (Freq/pat_total) * 100)
ovdata
```

Let us now display data for Frequency, stage, xray and operation (in that order) using **select()**
```{r}
select(ovdata, Freq, stage, xray, operation)
```
Lets use **filter()** now to just display the patients who survived. Then combine this code with a **pipe(%>%)** and a **select()** statement to display just the frequency and stage of those who survived.
```{r}
filter(ovdata, survival == "yes")

filter(ovdata, survival == "yes") %>%
  select(Freq,stage)
```
Now lets use **group_by()** and **summarise()** to produce a tibble like this (grouped by stage and xray and total number of patients in each combination).    
A tibble: 4 x 3   
   
Groups:   stage [2]    
stage    xray  total    
<fct>    <fct> <dbl>    
  1 early    no       65    
2 early    yes      93    
3 advanced no       48    
4 advanced yes      93    
```{r}
ovdata %>%
  group_by(stage, xray) %>%
  summarise(total = sum(Freq))
```
Let's now use **spread()** to reshape our data to display stage,operation,xray and survival-no, survival-yes as columns with the appropriate frequency in the last two columns like this:    
 tibble: 8 x 5    
  stage    operation xray     no   yes    
  <fct>    <fct>     <fct> <dbl> <dbl>    
1 early    radical   no       10    41    
2 early    radical   yes      17    64    
3 early    limited   no        1    13    
4 early    limited   yes       3     9    
5 advanced radical   no       38     6    
6 advanced radical   yes      64    11    
7 advanced limited   no        3     1    
8 advanced limited   yes      13     5    
```{r}
ovdata %>% spread(key=survival, value=Freq)
```

